---
title: "Relative Port Risk Summation"
output: html_document
---

# Purpose: To summarize the port and state data into a singular measure of relative risk per port
# Alternative: View in a multivariate approach like principal component analysis (PCA) or clustering to assess relative risk patterns across ports and states

# Summarize a single measure:
```{r combine-values}
library(here)
library(tidyverse)

state_to_port_final <- readRDS(here("data", "processed", "state_to_port_final.rds"))

relative_port_risk <- state_to_port_final |>
    mutate(
        relative_port_risk = 
        #percent of state exported; how many are going to each port
        (total_port_volume / total_state_bovine)
        *
        #total estimated at-risk cows; as a relative measure of the infestation level
        (detections_at_or_to / total_state_bovine)
        *
        #state at-risk cows
        (estimated_state_rate * total_state_bovine)
    ) |>
    group_by(port, year) |>
    summarise(
        relative_port_risk = sum(relative_port_risk, na.rm = TRUE),
        .groups = "drop"
    ) |>
    group_by(port) |>
    summarise(
        relative_port_risk = mean(relative_port_risk, na.rm = TRUE),
        .groups = "drop"
    ) |>
    arrange(desc(relative_port_risk))

total_chunks <- 10

relative_port_risk_chunks <- relative_port_risk |>
    mutate(
        risk_share = relative_port_risk / sum(relative_port_risk, na.rm = TRUE),
        raw_chunks = risk_share * total_chunks,
        chunks_floor = floor(raw_chunks),
        remainder = raw_chunks - chunks_floor
    ) |>
    arrange(desc(remainder), desc(relative_port_risk)) |>
    mutate(
        remainder_rank = row_number(),
        extra_chunks_needed = total_chunks - sum(chunks_floor),
        extra_chunk = as.integer(remainder_rank <= extra_chunks_needed),
        chunks_assigned = chunks_floor + extra_chunk
    ) |>
    select(port, relative_port_risk, risk_share, chunks_assigned) |>
    arrange(desc(chunks_assigned), desc(relative_port_risk))

glimpse(relative_port_risk)
View(relative_port_risk)
glimpse(relative_port_risk_chunks)
View(relative_port_risk_chunks)
```


# write relative port risk to output tables
```{r write-relative-port-risk}
saveRDS(relative_port_risk, here("data", "processed", "relative_port_risk.rds"))
write.csv(relative_port_risk, here("outputs", "tables", "relative_port_risk.csv"), row.names = FALSE)
saveRDS(relative_port_risk_chunks, here("data", "processed", "relative_port_risk_chunks.rds"))
write.csv(relative_port_risk_chunks, here("outputs", "tables", "relative_port_risk_chunks.csv"), row.names = FALSE)

```

# graph relative port risk
```{r graph-relative-port-risk}
library(ggplot2)

ggplot(
    relative_port_risk |>
        mutate(port = factor(port, levels = port[order(relative_port_risk, decreasing = TRUE)])),
    aes(x = port, y = relative_port_risk)
) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    labs(title = "Relative Port Risk", x = "Port", y = "Relative Risk")
```

# graph chunk assignment optimization
```{r graph-chunk-allocation}
chunk_plot_data <- relative_port_risk_chunks |>
    tidyr::uncount(chunks_assigned, .id = "chunk_number", .remove = FALSE) |>
    mutate(port = forcats::fct_reorder(port, chunks_assigned, .desc = TRUE))

ggplot(chunk_plot_data, aes(x = chunk_number, y = port, fill = risk_share)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_viridis_c(labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(breaks = 1:10) +
    theme_minimal() +
    labs(
        title = "10 Detector Dogs Allocated to Ports",
        subtitle = "Chunks assigned by largest-remainder rounding of each port's risk share",
        x = "# of dogs",
        y = "Port",
        fill = "Risk Share"
    )

# save to output figures
ggsave(here("outputs", "figures", "relative_port_risk_chunks.png"), width = 8, height = 6)

```