---
title: "Mexico Origins to Ports"
output: html_document
---

## PURPOSE: Estimate the risk when not all ports are open

# Method:

This notebook redistributes risk that was historically assigned to currently closed ports.

Primary model:
- Historical-share plus distance-decay choice model.

Sensitivity model:
- Gravity-style model.

Validation and stress tests:
- Fit on 2023 and predict 2024.
- Fit on 2024 and predict 2023.
- Leave-one-port-out substitution tests.

Primary rerouting outcome:
- at_risk_cattle_state_estimate

Sensitivity outcomes shown in rerouting outputs:
- nws_prevalence_from_state
- at_risk_cattle_port_estimate

```{r setup}
library(here)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(stringr)
library(ggplot2)
library(sf)

source("src/utils/parameters.R")
source("src/normalize_state_names.R")
source("src/assign_dogs.R")

set.seed(20260715)
```

```{r load-data}
state_to_port <- readRDS(here("data", "processed", "state_to_port_final.rds")) |>
	mutate(
		year = as.integer(year),
		port = str_to_lower(port),
		state_join = canonicalize_mexico_state(state)
	)

volume_df <- state_to_port |>
	dplyr::select(year, state_join, port, total_port_volume) |>
	mutate(total_port_volume = as.numeric(total_port_volume))

risk_outcomes <- c(
	"at_risk_cattle_state_estimate",
	"nws_prevalence_from_state",
	"at_risk_cattle_port_estimate"
)

years_available <- sort(unique(volume_df$year))
ports_available <- sort(unique(volume_df$port))

if (!all(opened_ports %in% ports_available)) {
	stop("One or more opened_ports are not present in state_to_port_final.")
}

if (!all(c(2023, 2024) %in% years_available)) {
	stop("This notebook expects 2023 and 2024 in state_to_port_final.")
}
```

```{r load-state-port-distance-table}
distance_candidates <- c(
	here("data", "processed", "state_port_drive_time.csv"),
	here("data", "processed", "state_port_distance_placeholder.csv")
)

distance_file <- distance_candidates[file.exists(distance_candidates)][1]

if (is.na(distance_file)) {
	stop(
		"No state-to-port distance table found. Run src/state_to_port_drive_time.R first."
	)
}

distance_placeholder <- read_csv(distance_file, show_col_types = FALSE) |>
	mutate(
		state_join = canonicalize_mexico_state(state_join),
		port = str_to_lower(port),
		distance_guess_km = as.numeric(distance_guess_km)
	) |>
	dplyr::select(state_join, port, distance_guess_km) |>
	distinct()

expected_pairs <- expand_grid(
	state_join = sort(unique(volume_df$state_join)),
	port = sort(unique(volume_df$port))
)

distance_placeholder <- expected_pairs |>
	left_join(distance_placeholder, by = c("state_join", "port"))

if (any(is.na(distance_placeholder$distance_guess_km))) {
	stop(
		"Distance table has missing values for one or more state/port pairs. ",
		"Regenerate with src/state_to_port_drive_time.R and ensure all routes succeed."
	)
}
```

```{r modeling-helpers}
add_state_totals <- function(df) {
	df |>
		group_by(year, state_join) |>
		mutate(total_state_volume = sum(total_port_volume, na.rm = TRUE)) |>
		ungroup()
}

add_history_from_reference <- function(df, ref_df) {
	hist_tbl <- ref_df |>
		group_by(state_join, port) |>
		summarise(hist_volume = sum(total_port_volume, na.rm = TRUE), .groups = "drop")

	df |>
		left_join(hist_tbl, by = c("state_join", "port")) |>
		mutate(hist_volume = coalesce(hist_volume, 0))
}

add_leave_year_out_history <- function(df) {
	pair_totals <- df |>
		group_by(state_join, port) |>
		summarise(total_pair_volume = sum(total_port_volume, na.rm = TRUE), .groups = "drop")

	df |>
		left_join(pair_totals, by = c("state_join", "port")) |>
		mutate(hist_volume = pmax(total_pair_volume - total_port_volume, 0)) |>
		dplyr::select(-total_pair_volume)
}

augment_model_inputs <- function(df, distance_tbl) {
	df |>
		left_join(distance_tbl, by = c("state_join", "port")) |>
		mutate(
			log_distance_km = log(pmax(distance_guess_km, 1)),
			log_hist_volume = log1p(hist_volume),
			total_state_volume = pmax(total_state_volume, 1)
		)
}

fit_primary_model <- function(train_df) {
	glm(
		total_port_volume ~ port + log_distance_km + log_hist_volume,
		offset = log(total_state_volume),
		family = quasipoisson(link = "log"),
		data = train_df
	)
}

fit_gravity_model <- function(train_df) {
	glm(
		total_port_volume ~ state_join + port + log_distance_km,
		offset = log(total_state_volume),
		family = quasipoisson(link = "log"),
		data = train_df
	)
}

predict_normalized_shares <- function(model_obj, new_df, open_port_subset = NULL) {
	d <- new_df
	if (!is.null(open_port_subset)) {
		d <- d |>
			filter(port %in% open_port_subset)
	}

	pred_vals <- as.numeric(predict(model_obj, newdata = d, type = "response"))

	d |>
		mutate(pred_flow = pmax(pred_vals, 0)) |>
		group_by(year, state_join) |>
		mutate(
			pred_share = {
				total_pred <- sum(pred_flow, na.rm = TRUE)
				if (total_pred > 0) {
					pred_flow / total_pred
				} else {
					rep(1 / n(), n())
				}
			}
		) |>
		ungroup()
}

evaluate_share_predictions <- function(pred_df, obs_df) {
	obs_shares <- obs_df |>
		group_by(year, state_join) |>
		mutate(obs_share = {
			total_obs <- sum(total_port_volume, na.rm = TRUE)
			if (total_obs > 0) {
				total_port_volume / total_obs
			} else {
				rep(0, n())
			}
		}) |>
		ungroup() |>
		dplyr::select(year, state_join, port, total_port_volume, obs_share)

	eval_df <- pred_df |>
		dplyr::select(year, state_join, port, pred_share, pred_flow) |>
		left_join(obs_shares, by = c("year", "state_join", "port")) |>
		mutate(
			abs_error_share = abs(pred_share - obs_share),
			sq_error_share = (pred_share - obs_share)^2,
			kl_term = if_else(
				obs_share > 0,
				obs_share * log(obs_share / pmax(pred_share, 1e-10)),
				0
			)
		)

	overall <- eval_df |>
		summarise(
			mae_share = mean(abs_error_share, na.rm = TRUE),
			rmse_share = sqrt(mean(sq_error_share, na.rm = TRUE)),
			mean_kl = mean(kl_term, na.rm = TRUE),
			.groups = "drop"
		)

	state_level <- eval_df |>
		group_by(year, state_join) |>
		summarise(
			mae_share = mean(abs_error_share, na.rm = TRUE),
			rmse_share = sqrt(mean(sq_error_share, na.rm = TRUE)),
			kl = sum(kl_term, na.rm = TRUE),
			observed_total_volume = sum(total_port_volume, na.rm = TRUE),
			.groups = "drop"
		)

	list(overall = overall, state_level = state_level, eval_df = eval_df)
}

run_cross_year <- function(volume_inputs, distance_tbl, train_year, test_year, method = c("primary", "gravity")) {
	method <- match.arg(method)

	train_df <- volume_inputs |>
		filter(year == train_year)

	test_df <- volume_inputs |>
		filter(year == test_year)

	train_df <- add_state_totals(train_df)
	test_df <- add_state_totals(test_df)

	train_df <- add_history_from_reference(train_df, train_df)
	test_df <- add_history_from_reference(test_df, train_df)

	train_df <- augment_model_inputs(train_df, distance_tbl)
	test_df <- augment_model_inputs(test_df, distance_tbl)

	model_obj <- if (method == "primary") {
		fit_primary_model(train_df)
	} else {
		fit_gravity_model(train_df)
	}

	predictions <- predict_normalized_shares(model_obj, test_df)
	eval <- evaluate_share_predictions(predictions, test_df)

	list(
		model = model_obj,
		predictions = predictions,
		overall = eval$overall |>
			mutate(train_year = train_year, test_year = test_year, method = method),
		state_level = eval$state_level |>
			mutate(train_year = train_year, test_year = test_year, method = method)
	)
}

run_leave_one_port_out <- function(
	volume_inputs,
	distance_tbl,
	train_year,
	test_year,
	closed_port,
	method = c("primary", "gravity")
) {
	method <- match.arg(method)

	train_df <- volume_inputs |>
		filter(year == train_year)

	test_df <- volume_inputs |>
		filter(year == test_year)

	train_df <- add_state_totals(train_df)
	test_df <- add_state_totals(test_df)

	train_df <- add_history_from_reference(train_df, train_df)
	test_df <- add_history_from_reference(test_df, train_df)

	train_df <- augment_model_inputs(train_df, distance_tbl)
	test_df <- augment_model_inputs(test_df, distance_tbl)

	model_obj <- if (method == "primary") {
		fit_primary_model(train_df)
	} else {
		fit_gravity_model(train_df)
	}

	states_with_closed_volume <- test_df |>
		filter(port == closed_port, total_port_volume > 0) |>
		distinct(state_join)

	if (nrow(states_with_closed_volume) == 0) {
		return(tibble(
			method = method,
			train_year = train_year,
			test_year = test_year,
			closed_port = closed_port,
			n_states = 0,
			closed_volume = 0,
			mae_share = NA_real_,
			rmse_share = NA_real_,
			mean_kl = NA_real_
		))
	}

	test_remaining <- test_df |>
		semi_join(states_with_closed_volume, by = "state_join") |>
		filter(port != closed_port)

	pred_remaining <- predict_normalized_shares(model_obj, test_remaining)

	obs_remaining <- test_remaining |>
		group_by(year, state_join) |>
		mutate(obs_share = {
			total_obs <- sum(total_port_volume, na.rm = TRUE)
			if (total_obs > 0) {
				total_port_volume / total_obs
			} else {
				rep(0, n())
			}
		}) |>
		ungroup() |>
		select(year, state_join, port, obs_share)

	eval_remaining <- pred_remaining |>
		dplyr::select(year, state_join, port, pred_share) |>
		dplyr::left_join(obs_remaining, by = c("year", "state_join", "port")) |>
		mutate(
			abs_error = abs(pred_share - obs_share),
			sq_error = (pred_share - obs_share)^2,
			kl_term = if_else(
				obs_share > 0,
				obs_share * log(obs_share / pmax(pred_share, 1e-10)),
				0
			)
		)

	closed_volume <- test_df |>
		semi_join(states_with_closed_volume, by = "state_join") |>
		filter(port == closed_port) |>
		summarise(v = sum(total_port_volume, na.rm = TRUE)) |>
		pull(v)

	tibble(
		method = method,
		train_year = train_year,
		test_year = test_year,
		closed_port = closed_port,
		n_states = n_distinct(eval_remaining$state_join),
		closed_volume = closed_volume,
		mae_share = mean(eval_remaining$abs_error, na.rm = TRUE),
		rmse_share = sqrt(mean(eval_remaining$sq_error, na.rm = TRUE)),
		mean_kl = mean(eval_remaining$kl_term, na.rm = TRUE)
	)
}

bootstrap_cross_year <- function(
	volume_inputs,
	distance_tbl,
	train_year,
	test_year,
	method = c("primary", "gravity"),
	n_boot = 100
) {
	method <- match.arg(method)

	train_df_full <- volume_inputs |>
		filter(year == train_year)
	test_df_full <- volume_inputs |>
		filter(year == test_year)

	state_ids <- sort(unique(train_df_full$state_join))

	boot_results <- vector("list", n_boot)

	for (b in seq_len(n_boot)) {
		sampled_states <- sample(state_ids, size = length(state_ids), replace = TRUE)
		state_weights <- as.data.frame(table(sampled_states), stringsAsFactors = FALSE)
		names(state_weights) <- c("state_join", "wt")

		train_df <- train_df_full |>
			left_join(state_weights, by = "state_join") |>
			mutate(wt = coalesce(as.numeric(wt), 0)) |>
			filter(wt > 0)

		test_df <- test_df_full |>
			left_join(state_weights, by = "state_join") |>
			mutate(wt = coalesce(as.numeric(wt), 0)) |>
			filter(wt > 0)

		train_df <- add_state_totals(train_df)
		test_df <- add_state_totals(test_df)

		train_df <- add_history_from_reference(train_df, train_df)
		test_df <- add_history_from_reference(test_df, train_df)

		train_df <- augment_model_inputs(train_df, distance_tbl)
		test_df <- augment_model_inputs(test_df, distance_tbl)

		model_obj <- if (method == "primary") {
			glm(
				total_port_volume ~ port + log_distance_km + log_hist_volume,
				offset = log(total_state_volume),
				family = quasipoisson(link = "log"),
				data = train_df,
				weights = wt
			)
		} else {
			glm(
				total_port_volume ~ state_join + port + log_distance_km,
				offset = log(total_state_volume),
				family = quasipoisson(link = "log"),
				data = train_df,
				weights = wt
			)
		}

		pred_df <- predict_normalized_shares(model_obj, test_df)
		eval <- evaluate_share_predictions(pred_df, test_df)

		boot_results[[b]] <- eval$overall |>
			mutate(
				method = method,
				train_year = train_year,
				test_year = test_year,
				boot_id = b
			)
	}

	bind_rows(boot_results)
}
```

```{r cross-year-validation}
cross_year_specs <- tibble::tribble(
	~train_year, ~test_year,
	2023L, 2024L,
	2024L, 2023L
)

cross_primary <- pmap(
	cross_year_specs,
	~run_cross_year(volume_df, distance_placeholder, ..1, ..2, method = "primary")
)

cross_gravity <- pmap(
	cross_year_specs,
	~run_cross_year(volume_df, distance_placeholder, ..1, ..2, method = "gravity")
)

cross_primary_overall <- bind_rows(lapply(cross_primary, function(x) x$overall))
cross_gravity_overall <- bind_rows(lapply(cross_gravity, function(x) x$overall))

cross_primary_state <- bind_rows(lapply(cross_primary, function(x) x$state_level))
cross_gravity_state <- bind_rows(lapply(cross_gravity, function(x) x$state_level))

write_csv(
	cross_primary_overall,
	here("outputs", "tables", "limited_port_crossyear_primary_overall.csv")
)

write_csv(
	cross_gravity_overall,
	here("outputs", "tables", "limited_port_crossyear_gravity_overall.csv")
)

write_csv(
	cross_primary_state,
	here("outputs", "tables", "limited_port_crossyear_primary_state.csv")
)

write_csv(
	cross_gravity_state,
	here("outputs", "tables", "limited_port_crossyear_gravity_state.csv")
)

cross_primary_overall
cross_gravity_overall
```

```{r leave-one-port-out-tests}
leave_specs <- expand_grid(
	method = c("primary", "gravity"),
	train_year = c(2023L, 2024L),
	test_year = c(2024L, 2023L),
	closed_port = ports_available
) |>
	filter(train_year != test_year)

leave_results <- pmap_dfr(
	leave_specs,
	function(method, train_year, test_year, closed_port) {
		run_leave_one_port_out(
			volume_inputs = volume_df,
			distance_tbl = distance_placeholder,
			train_year = train_year,
			test_year = test_year,
			closed_port = closed_port,
			method = method
		)
	}
)

write_csv(
	leave_results,
	here("outputs", "tables", "limited_port_leave_one_port_out_tests.csv")
)

leave_results
```

```{r bootstrap-uncertainty-validation}
n_boot <- 100

boot_validation <- bind_rows(
	bootstrap_cross_year(
		volume_inputs = volume_df,
		distance_tbl = distance_placeholder,
		train_year = 2023,
		test_year = 2024,
		method = "primary",
		n_boot = n_boot
	),
	bootstrap_cross_year(
		volume_inputs = volume_df,
		distance_tbl = distance_placeholder,
		train_year = 2024,
		test_year = 2023,
		method = "primary",
		n_boot = n_boot
	),
	bootstrap_cross_year(
		volume_inputs = volume_df,
		distance_tbl = distance_placeholder,
		train_year = 2023,
		test_year = 2024,
		method = "gravity",
		n_boot = n_boot
	),
	bootstrap_cross_year(
		volume_inputs = volume_df,
		distance_tbl = distance_placeholder,
		train_year = 2024,
		test_year = 2023,
		method = "gravity",
		n_boot = n_boot
	)
)

boot_summary <- boot_validation |>
	group_by(method, train_year, test_year) |>
	summarise(
		mae_share_p05 = quantile(mae_share, 0.05, na.rm = TRUE),
		mae_share_p50 = quantile(mae_share, 0.50, na.rm = TRUE),
		mae_share_p95 = quantile(mae_share, 0.95, na.rm = TRUE),
		rmse_share_p05 = quantile(rmse_share, 0.05, na.rm = TRUE),
		rmse_share_p50 = quantile(rmse_share, 0.50, na.rm = TRUE),
		rmse_share_p95 = quantile(rmse_share, 0.95, na.rm = TRUE),
		mean_kl_p05 = quantile(mean_kl, 0.05, na.rm = TRUE),
		mean_kl_p50 = quantile(mean_kl, 0.50, na.rm = TRUE),
		mean_kl_p95 = quantile(mean_kl, 0.95, na.rm = TRUE),
		.groups = "drop"
	)

write_csv(
	boot_validation,
	here("outputs", "tables", "limited_port_crossyear_bootstrap_draws.csv")
)

write_csv(
	boot_summary,
	here("outputs", "tables", "limited_port_crossyear_bootstrap_summary.csv")
)

boot_summary
```

```{r rerouting-model-fit-all-years}
scenario_volume <- volume_df |>
	add_state_totals() |>
	add_leave_year_out_history() |>
	augment_model_inputs(distance_placeholder)

primary_scenario_model <- fit_primary_model(scenario_volume)
gravity_scenario_model <- fit_gravity_model(scenario_volume)

scenario_probs_primary <- predict_normalized_shares(
	model_obj = primary_scenario_model,
	new_df = scenario_volume,
	open_port_subset = opened_ports
) |>
	dplyr::select(year, state_join, port, pred_share)

scenario_probs_gravity <- predict_normalized_shares(
	model_obj = gravity_scenario_model,
	new_df = scenario_volume,
	open_port_subset = opened_ports
) |>
	dplyr::select(year, state_join, port, pred_share)
```

```{r reroute-risk-primary-and-sensitivity}
risk_long <- state_to_port |>
	dplyr::select(year, state_join, port, all_of(risk_outcomes)) |>
	pivot_longer(cols = all_of(risk_outcomes), names_to = "outcome", values_to = "risk_value") |>
	mutate(risk_value = coalesce(as.numeric(risk_value), 0))

reroute_risk_with_probs <- function(risk_data_long, prob_tbl) {
	closed_risk <- risk_data_long |>
		filter(!port %in% opened_ports) |>
		group_by(year, state_join, outcome) |>
		summarise(closed_risk = sum(risk_value, na.rm = TRUE), .groups = "drop")

	existing_open <- risk_data_long |>
		filter(port %in% opened_ports) |>
		mutate(existing_open_risk = risk_value) |>
		dplyr::select(year, state_join, port, outcome, existing_open_risk)

	existing_open |>
		left_join(closed_risk, by = c("year", "state_join", "outcome")) |>
		left_join(prob_tbl, by = c("year", "state_join", "port")) |>
		mutate(
			closed_risk = coalesce(closed_risk, 0),
			pred_share = coalesce(pred_share, 0),
			rerouted_additional_risk = closed_risk * pred_share,
			rerouted_total_risk = existing_open_risk + rerouted_additional_risk
		)
}

rerouted_primary <- reroute_risk_with_probs(risk_long, scenario_probs_primary) |>
	mutate(model = "primary")

rerouted_gravity <- reroute_risk_with_probs(risk_long, scenario_probs_gravity) |>
	mutate(model = "gravity")

rerouted_all <- bind_rows(rerouted_primary, rerouted_gravity)

rerouted_port_year <- rerouted_primary |>
	group_by(year, port, outcome) |>
	summarise(
		rerouted_total_risk = sum(rerouted_total_risk, na.rm = TRUE),
		existing_open_risk = sum(existing_open_risk, na.rm = TRUE),
		rerouted_additional_risk = sum(rerouted_additional_risk, na.rm = TRUE),
		.groups = "drop"
	)

write_csv(
	rerouted_all,
	here("outputs", "tables", "limited_port_rerouted_state_port_risk.csv")
)

write_csv(
	rerouted_port_year,
	here("outputs", "tables", "limited_port_rerouted_port_year_risk_primary.csv")
)

rerouted_port_year |>
	filter(outcome == "at_risk_cattle_state_estimate")
```

```{r rerouting-uncertainty-primary-model}
scenario_open_rows <- scenario_volume |>
	filter(port %in% opened_ports)

beta_names <- colnames(vcov(primary_scenario_model))
X_all <- model.matrix(delete.response(terms(primary_scenario_model)), scenario_open_rows)
common_beta <- beta_names[beta_names %in% colnames(X_all)]
X_open <- X_all[, common_beta, drop = FALSE]
offset_open <- log(pmax(scenario_open_rows$total_state_volume, 1))
mu_beta <- coef(primary_scenario_model)[common_beta]
sigma_beta <- vcov(primary_scenario_model)[common_beta, common_beta, drop = FALSE]

coef_draws <- MASS::mvrnorm(
	n = 300,
	mu = mu_beta,
	Sigma = sigma_beta
)

closed_risk_by_state <- risk_long |>
	filter(!port %in% opened_ports) |>
	group_by(year, state_join, outcome) |>
	summarise(closed_risk = sum(risk_value, na.rm = TRUE), .groups = "drop")

existing_open_by_state <- risk_long |>
	filter(port %in% opened_ports) |>
	dplyr::select(year, state_join, port, outcome, existing_open_risk = risk_value)

draw_results <- vector("list", nrow(coef_draws))

for (i in seq_len(nrow(coef_draws))) {
	eta <- as.numeric(X_open %*% coef_draws[i, ]) + offset_open
	flow_draw <- exp(eta)

	prob_draw <- scenario_open_rows |>
		transmute(year, state_join, port, flow_draw) |>
		group_by(year, state_join) |>
		mutate(prob_draw = {
			total_draw <- sum(flow_draw, na.rm = TRUE)
			if (total_draw > 0) {
				flow_draw / total_draw
			} else {
				rep(1 / n(), n())
			}
		}) |>
		ungroup() |>
		dplyr::select(year, state_join, port, prob_draw)

	draw_tbl <- existing_open_by_state |>
		left_join(closed_risk_by_state, by = c("year", "state_join", "outcome")) |>
		left_join(prob_draw, by = c("year", "state_join", "port")) |>
		mutate(
			closed_risk = coalesce(closed_risk, 0),
			prob_draw = coalesce(prob_draw, 0),
			rerouted_total_risk = existing_open_risk + (closed_risk * prob_draw)
		) |>
		group_by(year, port, outcome) |>
		summarise(rerouted_total_risk = sum(rerouted_total_risk, na.rm = TRUE), .groups = "drop") |>
		mutate(draw_id = i)

	draw_results[[i]] <- draw_tbl
}

reroute_uncertainty_draws <- bind_rows(draw_results)

reroute_uncertainty_summary <- reroute_uncertainty_draws |>
	group_by(year, port, outcome) |>
	summarise(
		rerouted_total_risk_p05 = quantile(rerouted_total_risk, 0.05, na.rm = TRUE),
		rerouted_total_risk_p50 = quantile(rerouted_total_risk, 0.50, na.rm = TRUE),
		rerouted_total_risk_p95 = quantile(rerouted_total_risk, 0.95, na.rm = TRUE),
		.groups = "drop"
	)

write_csv(
	reroute_uncertainty_draws,
	here("outputs", "tables", "limited_port_rerouting_uncertainty_draws.csv")
)

write_csv(
	reroute_uncertainty_summary,
	here("outputs", "tables", "limited_port_rerouting_uncertainty_summary.csv")
)

reroute_uncertainty_summary |>
	filter(outcome == "at_risk_cattle_state_estimate")
```

```{r assign-dogs-rerouted-primary}
rerouted_for_chunks <- rerouted_port_year |>
	dplyr::select(year, port, outcome, rerouted_total_risk) |>
	tidyr::pivot_wider(names_from = outcome, values_from = rerouted_total_risk)

risk_chunk_results_rerouted <- assign_risk_chunks(
	data = rerouted_for_chunks,
	risk_columns = risk_outcomes,
	total_chunks = n_dogs,
	by_year = TRUE
)

dog_assignments_rerouted <- bind_rows(
	lapply(names(risk_chunk_results_rerouted), function(outcome_name) {
		risk_chunk_results_rerouted[[outcome_name]] |>
			mutate(outcome = outcome_name)
	})
)

write_csv(
	dog_assignments_rerouted,
	here("outputs", "tables", "limited_port_rerouted_dog_assignments.csv")
)

dog_assignments_rerouted |>
	filter(outcome == "at_risk_cattle_state_estimate") |>
	arrange(year, desc(chunks_assigned), desc(risk_value))
```

```{r assign-dogs-rerouted-uncertainty}
assign_dogs_for_risk_draws <- function(risk_draws, risk_columns, total_chunks) {
	draw_wide <- risk_draws |>
		dplyr::select(draw_id, year, port, outcome, rerouted_total_risk) |>
		tidyr::pivot_wider(names_from = outcome, values_from = rerouted_total_risk)

	draw_ids <- sort(unique(draw_wide$draw_id))

	purrr::map_dfr(draw_ids, function(current_draw_id) {
		draw_data <- draw_wide |>
			filter(draw_id == current_draw_id) |>
			dplyr::select(-draw_id)

		draw_chunk_results <- assign_risk_chunks(
			data = draw_data,
			risk_columns = risk_columns,
			total_chunks = total_chunks,
			by_year = TRUE
		)

		bind_rows(
			lapply(names(draw_chunk_results), function(outcome_name) {
				draw_chunk_results[[outcome_name]] |>
					mutate(
						draw_id = current_draw_id,
						outcome = outcome_name
					)
			})
		)
	})
}

dog_assignment_uncertainty_draws <- assign_dogs_for_risk_draws(
	risk_draws = reroute_uncertainty_draws,
	risk_columns = risk_outcomes,
	total_chunks = n_dogs
)

static_dog_assignments <- dog_assignments_rerouted |>
	dplyr::select(year, port, outcome, static_chunks_assigned = chunks_assigned)

dog_assignment_uncertainty_summary <- dog_assignment_uncertainty_draws |>
	left_join(static_dog_assignments, by = c("year", "port", "outcome")) |>
	group_by(year, port, outcome) |>
	summarise(
		dogs_min = min(chunks_assigned, na.rm = TRUE),
		dogs_p05 = as.integer(quantile(chunks_assigned, 0.05, na.rm = TRUE, type = 1)),
		dogs_p25 = as.integer(quantile(chunks_assigned, 0.25, na.rm = TRUE, type = 1)),
		dogs_p50 = as.integer(quantile(chunks_assigned, 0.50, na.rm = TRUE, type = 1)),
		dogs_p75 = as.integer(quantile(chunks_assigned, 0.75, na.rm = TRUE, type = 1)),
		dogs_p95 = as.integer(quantile(chunks_assigned, 0.95, na.rm = TRUE, type = 1)),
		dogs_max = max(chunks_assigned, na.rm = TRUE),
		dogs_mean = mean(chunks_assigned, na.rm = TRUE),
		prob_at_least_one_dog = mean(chunks_assigned >= 1, na.rm = TRUE),
		prob_at_least_static_assignment = mean(chunks_assigned >= first(static_chunks_assigned), na.rm = TRUE),
		static_chunks_assigned = first(static_chunks_assigned),
		.groups = "drop"
	) |>
	arrange(outcome, year, desc(dogs_p50), desc(dogs_p95), port)

write_csv(
	dog_assignment_uncertainty_draws,
	here("outputs", "tables", "limited_port_rerouted_dog_assignment_draws.csv")
)

write_csv(
	dog_assignment_uncertainty_summary,
	here("outputs", "tables", "limited_port_rerouted_dog_assignment_uncertainty.csv")
)

dog_assignment_uncertainty_summary |>
	filter(outcome == "at_risk_cattle_state_estimate")
```

```{r quick-figures, fig.width=10, fig.height=6}
plot_data <- rerouted_port_year |>
	filter(outcome == "at_risk_cattle_state_estimate")

ggplot(plot_data, aes(x = rerouted_total_risk, y = reorder(port, rerouted_total_risk), fill = factor(year))) +
	geom_col(position = "dodge") +
	labs(
		title = "Rerouted Port Risk Over Time (Primary Outcome)",
		x = "rerouted at-risk cattle (state estimate)",
		y = "port",
		fill = "year"
	) +
	theme_minimal(base_size = 12)

ggsave(
    here("outputs", "figures", "limited_port_rerouted_port_risk_primary.png"),
    width = 10,
    height = 6
)

dog_plot_data <- dog_assignments_rerouted |>
	filter(outcome == "at_risk_cattle_state_estimate")

ggplot(dog_plot_data, aes(x = chunks_assigned, y = reorder(port, chunks_assigned), fill = factor(year))) +
	geom_col(position = "dodge") +
	labs(
		title = "Rerouted Port Dog Assignments Over Time (Primary Outcome)",
		x = "Dogs assigned (state estimate)",
		y = "port",
		fill = "year"
	) +
	theme_minimal(base_size = 12)

ggsave(
    here("outputs", "figures", "limited_port_rerouted_port_dog_assignments_primary.png"),
    width = 10,
    height = 6
)

dog_uncertainty_plot_data <- dog_assignment_uncertainty_summary |>
	filter(outcome == "at_risk_cattle_state_estimate")

dog_uncertainty_position <- position_dodge(width = 0.55)

ggplot(
	dog_uncertainty_plot_data,
	aes(x = dogs_p50, y = reorder(port, dogs_p50), color = factor(year), group = factor(year))
) +
	geom_errorbarh(
		aes(xmin = dogs_p05, xmax = dogs_p95),
		height = 0.25,
		linewidth = 0.8,
		position = dog_uncertainty_position
	) +
    #plot the median assignment as a solid point
	geom_point(size = 2.5, position = dog_uncertainty_position) +
    #Plot the static assignment as an open circle
	geom_point(aes(x = static_chunks_assigned), shape = 1, size = 3, position = dog_uncertainty_position) +
	scale_x_continuous(breaks = seq(0, n_dogs, by = 1), limits = c(0, n_dogs)) +
	labs(
		title = "Rerouted Dog Assignment Uncertainty (Primary Outcome)",
		x = "dogs assigned: median with 5th to 95th percentile interval",
		y = "port",
		color = "year"
	) +
	theme_minimal(base_size = 12)

ggsave(
	here("outputs", "figures", "limited_port_rerouted_dog_assignment_uncertainty_primary.png"),
	width = 10,
	height = 6
)
```

```{r save-main-objects}
saveRDS(
	list(
		cross_primary = cross_primary,
		cross_gravity = cross_gravity,
		leave_one_port_out = leave_results,
		rerouted_primary = rerouted_primary,
		rerouted_gravity = rerouted_gravity,
		dog_assignments_rerouted = dog_assignments_rerouted,
		dog_assignment_uncertainty_draws = dog_assignment_uncertainty_draws,
		dog_assignment_uncertainty_summary = dog_assignment_uncertainty_summary,
		reroute_uncertainty_summary = reroute_uncertainty_summary
	),
	here("data", "processed", "limited_port_openings_model_outputs.rds")
)
```


