---
title: "NWS Spread in Mexico"
output: html_document
---

# Purpose: Explore the spread of NWS in Mexico, including mapping and analysis of rejection rates at ports.

## IDEAS
 # - Map detections in Mexico by location
 # - For any that are moving, map start and finish
    # - Also get the distance between start and finish
 # - Quantify how fast it is moving
    # - Like, a single polygon at time points, and the increase in area of the polygon over time?

## Notes:

 # - Human-assisted dispersal is primary method of spread
 # - No false positives; not sure the effectiveness of the surveys and the rate of false negatives
 # - The front is moving from central america; I do not know how many introductions there have been in Mexico

 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE
)
library(tidyverse)
library(readxl)
library(here)
#library(ggmap)
library(tidygeocoder)
library(sf)
library(stringi)
library(mapview)
```

# Load processed NWS detections data
```{r load-data}

nws_detections_geo <- readRDS(here("data", "processed", "nws_detections_geo.rds"))
glimpse(nws_detections_geo)

nws_detections_transport <- readRDS(here("data", "processed", "nws_detections_transport.rds"))
glimpse(nws_detections_transport)

```

## Map start and end locations of NWS detections
```{r map-start-end-locations}

required_cols <- c("origin_lon", "origin_lat", "destination_lon", "destination_lat")
missing_cols <- setdiff(required_cols, colnames(nws_detections_transport))
if (length(missing_cols) > 0) {
   stop(paste("Missing required columns in nws_detections_transport:", paste(missing_cols, collapse = ", ")))
}

mexico_states <- sf::st_read(
   here("data", "processed", "mexico_states_geo.geojson"),
   quiet = TRUE
) %>%
   sf::st_transform(4326)

origin_counts <- nws_detections_transport %>%
   dplyr::filter(!is.na(origin_lon), !is.na(origin_lat)) %>%
   dplyr::group_by(origin_lon, origin_lat) %>%
   dplyr::summarise(origin_count = dplyr::n(), .groups = "drop")

origin_points <- origin_counts %>%
   sf::st_as_sf(coords = c("origin_lon", "origin_lat"), crs = 4326, remove = FALSE)

destination_counts <- nws_detections_transport %>%
   dplyr::filter(!is.na(destination_lon), !is.na(destination_lat)) %>%
   dplyr::group_by(destination_lon, destination_lat) %>%
   dplyr::summarise(destination_count_point = dplyr::n(), .groups = "drop")

destination_points <- destination_counts %>%
   sf::st_as_sf(coords = c("destination_lon", "destination_lat"), crs = 4326, remove = FALSE)

# Keep non-aggregated destination records for state-level totals.
destination_records <- nws_detections_transport %>%
   dplyr::filter(!is.na(destination_lon), !is.na(destination_lat)) %>%
   sf::st_as_sf(coords = c("destination_lon", "destination_lat"), crs = 4326, remove = FALSE)

destination_with_state <- sf::st_join(
   destination_records,
   mexico_states %>% dplyr::select(CVEGEO),
   join = sf::st_intersects,
   left = TRUE
)

destination_counts_by_state <- destination_with_state %>%
   sf::st_drop_geometry() %>%
   dplyr::filter(!is.na(CVEGEO)) %>%
   dplyr::count(CVEGEO, name = "destination_count")

states_destination_map <- mexico_states %>%
   dplyr::left_join(destination_counts_by_state, by = "CVEGEO") %>%
   dplyr::mutate(destination_count = tidyr::replace_na(destination_count, 0L))

flow_counts <- nws_detections_transport %>%
   dplyr::filter(
      !is.na(origin_lon),
      !is.na(origin_lat),
      !is.na(destination_lon),
      !is.na(destination_lat)
   ) %>%
   dplyr::group_by(origin_lon, origin_lat, destination_lon, destination_lat) %>%
   dplyr::summarise(n_flow = dplyr::n(), .groups = "drop") %>%
   dplyr::mutate(flow_id = dplyr::row_number())

if (nrow(flow_counts) == 0) {
   stop("No complete origin/destination coordinates found to build flow lines.")
}

flow_lines <- sf::st_sf(
   flow_id = flow_counts$flow_id,
   n_flow = flow_counts$n_flow,
   geometry = sf::st_sfc(
      lapply(seq_len(nrow(flow_counts)), function(i) {
         sf::st_linestring(
            matrix(
               c(
                  flow_counts$origin_lon[i], flow_counts$origin_lat[i],
                  flow_counts$destination_lon[i], flow_counts$destination_lat[i]
               ),
               ncol = 2,
               byrow = TRUE
            )
         )
      }),
      crs = 4326
   )
)

destination_flow_points <- flow_counts %>%
   sf::st_as_sf(coords = c("destination_lon", "destination_lat"), crs = 4326, remove = FALSE)

# Build geometry-based arrowheads (no external arrow package required).
flow_coords_3857 <- flow_counts %>%
   dplyr::select(flow_id, n_flow, origin_lon, origin_lat, destination_lon, destination_lat) %>%
   sf::st_as_sf(coords = c("origin_lon", "origin_lat"), crs = 4326, remove = FALSE) %>%
   sf::st_transform(3857)

origin_xy <- sf::st_coordinates(flow_coords_3857)

dest_coords_3857 <- flow_counts %>%
   sf::st_as_sf(coords = c("destination_lon", "destination_lat"), crs = 4326, remove = FALSE) %>%
   sf::st_transform(3857)

dest_xy <- sf::st_coordinates(dest_coords_3857)

rotate_vec <- function(vx, vy, angle_rad) {
   c(
      vx * cos(angle_rad) - vy * sin(angle_rad),
      vx * sin(angle_rad) + vy * cos(angle_rad)
   )
}

arrow_size <- scales::rescale(log1p(flow_counts$n_flow), to = c(18000, 52000))
arrow_angle <- pi / 7

arrow_rows <- lapply(seq_len(nrow(flow_counts)), function(i) {
   vec <- c(dest_xy[i, 1] - origin_xy[i, 1], dest_xy[i, 2] - origin_xy[i, 2])
   vec_norm <- sqrt(sum(vec^2))

   if (is.na(vec_norm) || vec_norm == 0) {
      return(NULL)
   }

   u <- vec / vec_norm
   wing1 <- rotate_vec(u[1], u[2], pi - arrow_angle)
   wing2 <- rotate_vec(u[1], u[2], pi + arrow_angle)

   tip <- dest_xy[i, ]
   base1 <- tip + arrow_size[i] * wing1
   base2 <- tip + arrow_size[i] * wing2

   tibble::tibble(
      flow_id = flow_counts$flow_id[i],
      n_flow = flow_counts$n_flow[i],
      geom1 = list(sf::st_linestring(rbind(tip, base1))),
      geom2 = list(sf::st_linestring(rbind(tip, base2)))
   )
})

arrow_table <- dplyr::bind_rows(arrow_rows)

arrowhead_segments <- dplyr::bind_rows(
   arrow_table %>% dplyr::transmute(flow_id, n_flow, geometry = geom1),
   arrow_table %>% dplyr::transmute(flow_id, n_flow, geometry = geom2)
) %>%
   sf::st_as_sf(crs = 3857) %>%
   sf::st_transform(4326)

# Repeat each unique line by its count so overlap darkens naturally via alpha blending.
line_repeat_index <- rep(seq_len(nrow(flow_lines)), flow_lines$n_flow)
flow_lines_alpha <- flow_lines[line_repeat_index, ]

cat("Number of unique origin-destination flow lines:", nrow(flow_lines), "\n")
cat("Number of transport records rendered as alpha-stacked lines:", nrow(flow_lines_alpha), "\n")
cat("Number of unique origin points:", nrow(origin_points), "\n")
cat("Number of unique destination points:", nrow(destination_points), "\n")

flow_map <- mapview::mapview(
   states_destination_map,
   zcol = "destination_count",
   layer.name = "Destinations by state",
   lwd = 1,
   alpha.regions = 0.75,
   col.regions = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(8)
) +
   mapview::mapview(flow_lines_alpha, color = "firebrick", lwd = 2.5, alpha = 0.08, layer.name = "Spread path (alpha overlap)", legend = FALSE) +
   mapview::mapview(arrowhead_segments, color = "firebrick", lwd = 2.8, alpha = 1, layer.name = "Flow direction") +
   mapview::mapview(
      origin_points,
      zcol = "origin_count",
      col.regions = colorRampPalette(c("#c6dbef", "#2171b5", "#08306b"))(8),
      cex = 5,
      alpha = 0.95,
      layer.name = "Origin (count at point)"
   ) +
   mapview::mapview(
      destination_points,
      zcol = "destination_count_point",
      col.regions = colorRampPalette(c("#fdd0a2", "#e6550d", "#7f2704"))(8),
      cex = 5,
      alpha = 0.95,
      layer.name = "Destination (count at point)"
   )

output_dir <- here("outputs", "figures")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

htmlwidgets::saveWidget(
   flow_map@map,
   file = file.path(output_dir, "flow_map.html"),
   selfcontained = TRUE
)

cat("Saved interactive map to:", file.path(output_dir, "flow_map.html"), "\n")

flow_map



```

## Map detections by state
```{r map-detections-by-state}

mexico_states_geo <- sf::st_read(
   here("data", "processed", "mexico_states_geo.geojson"),
   quiet = TRUE
) %>%
   sf::st_transform(4326)

detection_points <- nws_detections_geo %>%
   sf::st_make_valid() %>%
   sf::st_transform(4326) %>%
   dplyr::filter(!sf::st_is_empty(.)) %>%
   sf::st_point_on_surface()

if ("State" %in% colnames(nws_detections_geo)) {
   detections_by_state <- nws_detections_geo %>%
      sf::st_drop_geometry() %>%
      dplyr::filter(!is.na(State)) %>%
      dplyr::count(State, name = "detection_count")

   states_detection_map <- mexico_states_geo %>%
      dplyr::left_join(detections_by_state, by = c("NOMGEO" = "State")) %>%
      dplyr::mutate(detection_count = tidyr::replace_na(detection_count, 0L))
} else {
   detections_with_state <- sf::st_join(
      detection_points,
      mexico_states_geo %>% dplyr::select(CVEGEO),
      join = sf::st_intersects,
      left = TRUE
   )

   detections_by_state <- detections_with_state %>%
      sf::st_drop_geometry() %>%
      dplyr::filter(!is.na(CVEGEO)) %>%
      dplyr::count(CVEGEO, name = "detection_count")

   states_detection_map <- mexico_states_geo %>%
      dplyr::left_join(detections_by_state, by = "CVEGEO") %>%
      dplyr::mutate(detection_count = tidyr::replace_na(detection_count, 0L))
}

cat("States with at least one detection:", sum(states_detection_map$detection_count > 0), "\n")

states_detection_map <- states_detection_map %>%
   dplyr::mutate(
      area_km2 = as.numeric(sf::st_area(sf::st_transform(geometry, 6933))) / 1e6,
      detection_density = dplyr::if_else(area_km2 > 0, detection_count / area_km2, NA_real_)
   )

cat("Maximum state detection density (detections per km^2):", round(max(states_detection_map$detection_density, na.rm = TRUE), 4), "\n")

state_map <- mapview::mapview(
   states_detection_map,
   zcol = "detection_density",
   layer.name = "State detection density",
   lwd = 0,
   alpha.regions = 0.85,
   col.regions = colorRampPalette(c("#f7fbff", "#6baed6", "#08306b"))(8)
)

output_dir <- here("outputs", "figures")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

htmlwidgets::saveWidget(
   state_map@map,
   file = file.path(output_dir, "state_detections_map.html"),
   selfcontained = TRUE
)

cat("Saved interactive state map to:", file.path(output_dir, "state_detections_map.html"), "\n")

state_map

```

## Heatmap of all detection locations
```{r map-all-detections-heatmap}

# Build detection points robustly whether nws_detections_geo is sf or a plain data frame.
if (inherits(nws_detections_geo, "sf")) {
   detection_points_all <- nws_detections_geo %>%
      sf::st_make_valid() %>%
      sf::st_transform(4326) %>%
      dplyr::filter(!sf::st_is_empty(.)) %>%
      sf::st_point_on_surface()

   detection_coords <- sf::st_coordinates(detection_points_all)

   nws_detection_points_latlon <- detection_points_all %>%
      sf::st_drop_geometry() %>%
      dplyr::mutate(
         lon = detection_coords[, 1],
         lat = detection_coords[, 2]
      )
} else {
   geo_col_candidates <- c("municipality_geometry", "state_geometry", "geometry")
   geo_col_candidates <- geo_col_candidates[geo_col_candidates %in% colnames(nws_detections_geo)]

   geo_col <- NA_character_
   if (length(geo_col_candidates) > 0) {
      for (cand in geo_col_candidates) {
         if (inherits(nws_detections_geo[[cand]], "sfc")) {
            geo_col <- cand
            break
         }
      }
   }

   if (!is.na(geo_col)) {
      detection_points_all <- nws_detections_geo %>%
         sf::st_as_sf(sf_column_name = geo_col) %>%
         sf::st_make_valid() %>%
         sf::st_transform(4326) %>%
         dplyr::filter(!sf::st_is_empty(.)) %>%
         sf::st_point_on_surface()

      detection_coords <- sf::st_coordinates(detection_points_all)

      nws_detection_points_latlon <- detection_points_all %>%
         sf::st_drop_geometry() %>%
         dplyr::mutate(
            lon = detection_coords[, 1],
            lat = detection_coords[, 2]
         )
   } else {
      lon_candidates <- c("lon", "Lon", "LON", "longitude", "Longitude", "LONGITUDE", "x", "X")
      lat_candidates <- c("lat", "Lat", "LAT", "latitude", "Latitude", "LATITUDE", "y", "Y")

      lon_col <- lon_candidates[lon_candidates %in% colnames(nws_detections_geo)][1]
      lat_col <- lat_candidates[lat_candidates %in% colnames(nws_detections_geo)][1]

      if (is.na(lon_col) || is.na(lat_col)) {
         stop(
            "Could not find geometry or lon/lat columns in nws_detections_geo. ",
            "Expected an sf object, an sfc column (e.g., municipality_geometry), or lon/lat columns."
         )
      }

      nws_detection_points_latlon <- nws_detections_geo %>%
         dplyr::mutate(
            lon = .data[[lon_col]],
            lat = .data[[lat_col]]
         ) %>%
         dplyr::filter(!is.na(lon), !is.na(lat))

      detection_points_all <- nws_detection_points_latlon %>%
         sf::st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
   }
}

cat("Total detections with lat/lon:", nrow(nws_detection_points_latlon), "\n")

# Destination points come from transport records.
destination_points_latlon <- nws_detections_transport %>%
   dplyr::filter(!is.na(destination_lon), !is.na(destination_lat)) %>%
   dplyr::transmute(
      lon = destination_lon,
      lat = destination_lat
   )

cat("Total destinations with lat/lon:", nrow(destination_points_latlon), "\n")

output_dir <- here("outputs", "figures")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

if (requireNamespace("leaflet.extras", quietly = TRUE)) {
   detection_heatmap <- leaflet::leaflet() %>%
      leaflet::addProviderTiles("CartoDB.Positron") %>%
      leaflet.extras::addHeatmap(
         data = nws_detection_points_latlon,
         lng = ~lon,
         lat = ~lat,
         blur = 18,
         radius = 14,
         max = 1,
         minOpacity = 0.35,
         group = "Observations",
         gradient = c("0.4" = "#c6dbef", "0.65" = "#4292c6", "1.0" = "#08306b")
      )

   if (nrow(destination_points_latlon) > 0) {
      detection_heatmap <- detection_heatmap %>%
         leaflet.extras::addHeatmap(
            data = destination_points_latlon,
            lng = ~lon,
            lat = ~lat,
            blur = 18,
            radius = 14,
            max = 1,
            minOpacity = 0.35,
            group = "Destinations",
            gradient = c("0.4" = "#fdd0a2", "0.65" = "#f16913", "1.0" = "#7f2704")
         ) %>%
         leaflet::addLayersControl(
            overlayGroups = c("Observations", "Destinations"),
            options = leaflet::layersControlOptions(collapsed = FALSE)
         )
   } else {
      message("No destination coordinates found; only observation heatmap layer was added.")
      detection_heatmap <- detection_heatmap %>%
         leaflet::addLayersControl(
            overlayGroups = c("Observations"),
            options = leaflet::layersControlOptions(collapsed = FALSE)
         )
   }

   htmlwidgets::saveWidget(
      detection_heatmap,
      file = file.path(output_dir, "nws_detections_heatmap.html"),
      selfcontained = TRUE
   )

   cat("Saved interactive heatmap to:", file.path(output_dir, "nws_detections_heatmap.html"), "\n")
   detection_heatmap
} else {
   message("leaflet.extras not installed; using hex-density fallback map.")

   detections_3857 <- sf::st_transform(detection_points_all, 3857)
   destinations_3857 <- destination_points_latlon %>%
      sf::st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE) %>%
      sf::st_transform(3857)

   mexico_states_3857 <- sf::st_transform(mexico_states_geo, 3857)

   heat_grid <- sf::st_make_grid(
      sf::st_union(mexico_states_3857),
      cellsize = 50000,
      square = FALSE
   ) %>%
      sf::st_as_sf() %>%
      dplyr::mutate(
         observation_count = lengths(sf::st_intersects(., detections_3857)),
         destination_count = lengths(sf::st_intersects(., destinations_3857))
      ) %>%
      dplyr::filter(observation_count > 0 | destination_count > 0) %>%
      sf::st_transform(4326)

   observation_hex <- heat_grid %>%
      dplyr::filter(observation_count > 0)

   destination_hex <- heat_grid %>%
      dplyr::filter(destination_count > 0)

   detection_heatmap <- mapview::mapview(
      observation_hex,
      zcol = "observation_count",
      layer.name = "Observation heatmap (hex)",
      lwd = 0,
      alpha.regions = 0.7,
      col.regions = colorRampPalette(c("#deebf7", "#4292c6", "#084594"))(8)
   )

   if (nrow(destination_hex) > 0) {
      detection_heatmap <- detection_heatmap +
         mapview::mapview(
            destination_hex,
            zcol = "destination_count",
            layer.name = "Destination heatmap (hex)",
            lwd = 0,
            alpha.regions = 0.55,
            col.regions = colorRampPalette(c("#fee6ce", "#f16913", "#7f2704"))(8)
         )
   } else {
      message("No destination coordinates found; only observation hex layer was added.")
   }

   htmlwidgets::saveWidget(
      detection_heatmap@map,
      file = file.path(output_dir, "nws_detections_heatmap.html"),
      selfcontained = TRUE
   )

   cat("Saved interactive heatmap to:", file.path(output_dir, "nws_detections_heatmap.html"), "\n")
   detection_heatmap
}

```
