---
title: "Mexico Origins to Ports"
output: html_document
---

# Purpose: Explore the data on ports and Mexico state origins.


 - going to need the bovines per municipality data
    - Needs to be geolocated as well
 - imports by mexican state and port per year

# Quantify:
    - Connections (volume of import) between state and port
    - Percent of state inventory moving to each port
    - Proportions of detections relative to state inventory
    - Proportion of infested transports to state inventory


# setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(sf)
library(stringi)
```

# Load georeferenced bovine data by state
```{r summarize-bovine-by-state}
bovine_state_summary <- readRDS(here("data", "processed", "geolocated_bovine_state_summary.rds"))
glimpse(bovine_state_summary)
```

# Load port data

```{r load-port-data}
port_volume_by_state <- readRDS(here("data", "processed", "port_volume_by_state.rds"))
glimpse(port_volume_by_state)
```


#    - Percent of state inventory moving to each port
```{r percent-state-inventory-to-port}
normalize_state_name <- function(x) {
    x |>
        stringi::stri_trans_general("Latin-ASCII") |>
        stringr::str_to_lower() |>
        stringr::str_replace_all("[^a-z0-9]+", " ") |>
        stringr::str_squish()
}

## Align state names between the two processed datasets.
#state_name_overrides <- c(
#    "michioacan" = "michoacan",
#    "distrito federal" = "ciudad de mexico",
#    "mexico" = "estado de mexico"
#)

port_state_clean <- port_volume_by_state |>
    mutate(
        state_norm = normalize_state_name(state)
        #,
        #state_norm = recode(state_norm, !!!state_name_overrides)
    )

bovine_state_clean <- bovine_state_summary |>
    sf::st_drop_geometry() |>
    transmute(
        state = municipio,
        total_state_bovine = total_de_bovinos,
        state_norm = normalize_state_name(municipio)
    )

state_port_joined <- port_state_clean |>
    left_join(
        bovine_state_clean,
        by = "state_norm",
        relationship = "many-to-one"
    ) |> rename(total_port_volume = total_volume)

unmatched_states <- state_port_joined |>
    filter(is.na(total_state_bovine)) |>
    distinct(state = state.x, state_norm) |>
    arrange(state)

unmatched_states

# 1) Total volume from each state to each port.
state_to_port_total_volume <- state_port_joined |>
    transmute(
        year,
        state = coalesce(state.y, state.x),
        port,
        total_port_volume
    ) |>
    arrange(year, state, port)

state_to_port_total_volume_wide <- state_to_port_total_volume |>
    pivot_wider(
        names_from = port,
        values_from = total_port_volume,
        values_fill = 0
    ) |>
    arrange(year, state)

state_to_port_total_volume_wide

# 2) Total volume from each state to each port divided by total bovine in each state.
state_to_port_volume_per_bovine <- state_port_joined |>
    transmute(
        year,
        state = coalesce(state.y, state.x),
        port,
        total_port_volume,
        total_state_bovine,
        port_volume_per_bovine = if_else(total_state_bovine > 0, total_port_volume / total_state_bovine, NA_real_)
    ) |>
    arrange(year, state, port)

state_to_port_volume_per_bovine_wide <- state_to_port_volume_per_bovine |>
    select(year, state, port, port_volume_per_bovine) |>
    pivot_wider(
        names_from = port,
        values_from = port_volume_per_bovine,
        values_fill = 0
    ) |>
    arrange(year, state)

state_to_port_volume_per_bovine_wide

write.csv(
    state_to_port_total_volume,
    here("outputs", "tables", "state_to_port_total_volume.csv"),
    row.names = FALSE
)

write.csv(
    state_to_port_volume_per_bovine,
    here("outputs", "tables", "state_to_port_volume_per_bovine.csv"),
    row.names = FALSE
)


print("Come back to these that are not matched!!:")
print(unmatched_states)

```

# Combine the at-risk rate per port

```{r combine-at-risk-rate-per-port}
# rejection_rates <- readRDS(here("data", "processed", "mean_rejection_rates_with_tukey.rds"))
# glimpse(rejection_rates)

# state_to_port_tmp <- state_to_port_volume_per_bovine |> 
# dplyr::left_join(rejection_rates, by = "port")

# explore using rejection rate by year instead
rejection_rates_yr <- readRDS(here("data", "processed", "rejection_rates.rds"))
rejection_rates_yr <- rejection_rates_yr |> 
    mutate(year = as.numeric(as.character(year)))
state_to_port_tmp <- state_to_port_volume_per_bovine |> 
dplyr::left_join(rejection_rates_yr, by = c("port", "year"))

glimpse(state_to_port_tmp)
```

```{r add-detections}
detections <- readRDS(here("data", "processed", "nws_detections.rds"))
glimpse(detections)


#summarize detections by state
detections_by_state <- detections |>
    group_by(State) |>
    summarise(total_detections = n())

glimpse(detections_by_state)

state_to_port_tmp <- state_to_port_tmp |>
    left_join(detections_by_state, by = c("state" = "State"))

glimpse(state_to_port_tmp)

state_to_port_tmp <- state_to_port_tmp |>
    mutate(total_detections = replace_na(total_detections, 0))

glimpse(state_to_port_tmp)
```

```{r add destination to detections}
# Detection data doesn't account for movement. Add the destination of the detection to the state
transport <- read.csv(
    here("data", "processed", "nws_detections_transport.csv")
)
glimpse(transport)

transport <- transport |>
    group_by(destination_matched_state) |>
    summarise(n_traveling_to = n())

state_to_port_tmp <- state_to_port_tmp |>
    left_join(transport, by = c("state" = "destination_matched_state")) |>
    mutate(n_traveling_to = replace_na(n_traveling_to, 0))

state_to_port_tmp <- state_to_port_tmp |> 
    mutate(detections_at_or_to = total_detections + n_traveling_to)

glimpse(state_to_port_tmp)
```

# use proportional allocation to determine the rate of at-risk bovine in each area


# Calculate the proportion of each port's volume by state
 - Turn it into a matrix and calcualte the proportion of the sum

 ```{r proportion of port volume by state}
state_prop <- state_to_port_tmp |>
    dplyr::select(year, state, port, total_port_volume) |>
    ungroup() |>
    pivot_wider(
        names_from = state,
        values_from = total_port_volume
    )

state_prop <- state_prop |>
    mutate(across(-c(year, port), ~ .x / sum(.x, na.rm = TRUE))) |>
    mutate(across(-c(year, port), ~ replace_na(.x, 0)))

# pivot_longer and match
state_prop <- state_prop |>
    pivot_longer(
        cols = -c(year, port),
        names_to = "state",
        values_to = "proportion_port_volume"
    )

state_to_port_tmp2 <- state_to_port_tmp |>
    left_join(state_prop, by = c("year", "port", "state"))
glimpse(state_to_port_tmp2)
View(state_to_port_tmp2)
 ```

# calculate the estimated state-wide at-risk bovine
```{r rejection-rate-by-state}
state_rate <- state_to_port_tmp2 |>
    mutate(estimated_state_rate = proportion_port_volume * rejection_rate) |>
    group_by(state, year) |>
    summarise(estimated_state_rate = sum(estimated_state_rate, na.rm = TRUE))

glimpse(state_rate)
View(state_rate)
```



# Formalize the final dataset
```{r formalize-final-dataset}
state_to_port_tmp2 <- state_to_port_tmp2 |> left_join(state_rate, by = c("state", "year"))


state_to_port_final <- state_to_port_tmp2
glimpse(state_to_port_final)
View(state_to_port_final)

# Write to final output tables
saveRDS(
    state_to_port_final,
    here("outputs", "tables", "state_to_port_final.rds")
)

write.csv(
    state_to_port_final,
    here("outputs", "tables", "state_to_port_final.csv"),
    row.names = FALSE
)

# Write to processed data folder

saveRDS(
    state_to_port_final,
    here("data", "processed", "state_to_port_final.rds")
)
write.csv(
    state_to_port_final,
    here("data", "processed", "state_to_port_final.csv"),
    row.names = FALSE
)
```
