---
title: "NWS Spread in Mexico"
output: html_document
---

# Purpose: Explore the spread of NWS in Mexico, including mapping and analysis of rejection rates at ports.

## IDEAS
 # - Map detections in Mexico by location
 # - For any that are moving, map start and finish
    # - Also get the distance between start and finish
 # - Quantify how fast it is moving
    # - Like, a single polygon at time points, and the increase in area of the polygon over time?

## Notes:

 # - Human-assisted dispersal is primary method of spread
 # - No false positives; not sure the effectiveness of the surveys and the rate of false negatives
 # - The front is moving from central america; I do not know how many introductions there have been in Mexico

 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE
)
library(tidyverse)
library(readxl)
library(here)
#library(ggmap)
library(tidygeocoder)
library(sf)
library(stringi)
```

# Load processed NWS detections data
```{r load-data}

nws_detections <- readRDS(here("data", "processed", "nws_detections.rds"))
glimpse(nws_detections)


```






