---
title: "Mexico Origins to Ports"
output: html_document
---

## PURPOSE: Estimate the rejection rate of each state in Mexico 
- based on the proportion of each port's volume and the rejection rate of that port
- this will be used to estimate the number of at-risk cattle in each state that are moving to the US
- this assumes that there are not differences in rejection rates in cattle lots going from the same state but to different ports of entry
    -- This is a VERY strong assumption and likely not true

HOWEVER, we see some clear port profiles in terms of rejection rates, so possibly there is some truth to it


NOTE -- just thinking... : This assumption could be tested by considering the distance between rejection profiles of ports and the amount of overlap from same-state sourcing

```{r load-libraries}
library(dplyr)
library(here)
library(ggplot2)
```


```{r load state to port volume per bovine} 
state_to_port_volume_per_bovine <- read.csv(
    here("outputs", "tables", "state_to_port_volume_per_bovine.csv"),
)
```

# Combine the at-risk rate per port

```{r combine-at-risk-rate-per-port}
# explore using rejection rate by year instead
rejection_rates_yr <- readRDS(here("data", "processed", "rejection_rates.rds"))
rejection_rates_yr <- rejection_rates_yr |> 
    mutate(year = as.numeric(as.character(year)))
state_to_port_tmp <- state_to_port_volume_per_bovine |> 
dplyr::left_join(rejection_rates_yr, by = c("port", "year"))

glimpse(state_to_port_tmp)
```

# use proportional allocation to determine the rate of at-risk bovine in each area
# Calculate the proportion of each port's volume by state
 - Turn it into a matrix and calcualte the proportion of the sum

 ```{r proportion of port volume by state}
state_prop <- state_to_port_tmp |>
    dplyr::select(year, state, port, total_port_volume)

state_prop <- state_to_port_tmp |>
    dplyr::select(year, state, port, total_port_volume) |>
    group_by(year, port) |>
    mutate(proportion_port_volume = total_port_volume / sum(total_port_volume, na.rm = TRUE)) |>
    ungroup()

#sanity check:
sanity_check <- state_prop |> group_by(port,year) |> summarise(total_proportion_port_volume = sum(proportion_port_volume))
if (any(is.na(sanity_check$total_proportion_port_volume))) {
    warning("There are NA values in the total_proportion_port_volume column.")
} else if (any(sanity_check$total_proportion_port_volume != 0 & sanity_check$total_proportion_port_volume != 1)) {
    warning("There are values outside 0 and 1 in the total_proportion_port_volume column.")
} else {
    cat("All values in total_proportion_port_volume are either 1 or 0 [0, 1], as intended.\n See object sanity_check in code")
}

state_to_port_tmp2 <- state_to_port_tmp |>
    left_join(state_prop, by = c("year", "port", "state"))
glimpse(state_to_port_tmp2)
#View(state_to_port_tmp2)
 ```

# calculate the estimated state-wide at-risk bovine
```{r rejection-rate-by-state}
state_rate <- state_to_port_tmp |>
  group_by(state, year) |>
  summarise(
    total_state_shipped = sum(total_port_volume, na.rm = TRUE),
    estimated_at_risk_cattle = sum(total_port_volume * rejection_rate, na.rm = TRUE),
    estimated_state_rate_shipped = if_else(
      total_state_shipped > 0,
      estimated_at_risk_cattle / total_state_shipped,
      0,
      missing = 0
    ),
    estimated_state_rate_weighted = if_else(
      total_state_shipped > 0,
      sum((total_port_volume / total_state_shipped) * rejection_rate, na.rm = TRUE),
      0,
      missing = 0
    ),
    .groups = "drop"
  ) |>
  mutate(
    across(
      c(estimated_state_rate_shipped, estimated_state_rate_weighted),
      ~ tidyr::replace_na(.x, 0)
    )
  )


# #estimate the rejection rate of each stateXyear based on the proportion of each port's volume and the rejection rate of that port 
# state_rate <- state_to_port_tmp2 |>
#     mutate(estimated_state_rate = proportion_port_volume * rejection_rate) |>
#     group_by(state, year) |>
#     mutate(total_state_shipped = sum(total_port_volume, na.rm = TRUE)) |>
#     summarise(estimated_state_rate = sum(estimated_state_rate, na.rm = TRUE))

glimpse(state_rate)
#View(state_rate)

# make graph to visualize the yearly estimates
library(ggplot2)
state_rate_plot <- ggplot(state_rate, aes(x = year, y = estimated_state_rate_weighted, color = state)) +
    geom_line() +
    geom_point() +
    theme_minimal() +
    labs(
        title = "Estimated State Rejection Rates Over Time",
        x = "Year",
        y = "Estimated State Rejection Rate"
    )

state_rate_plot

state_rate_plot2 <- ggplot(state_rate, aes(x = year, y = estimated_state_rate_shipped, color = state)) +
    geom_line() +
    geom_point() +
    theme_minimal() +
    labs(
        title = "Estimated State Rejection Rates Over Time",
        x = "Year",
        y = "Estimated State Rejection Rate"
    )

state_rate_plot2
```

# Save
```{r save-data-and-graphs}
# save state_rate estimates
saveRDS(
    state_rate,
    here("data", "processed", "state_rate.rds")
)
write.csv(
    state_rate,
    here("data", "processed", "state_rate.csv"),
    row.names = FALSE
)
write.csv(
    state_rate,
    here("outputs", "tables", "state_rate.csv"),
    row.names = FALSE
)

# save state_rate graph
ggsave(
    here("outputs", "figures", "state_rate_plot.png"),
    plot = state_rate_plot,
    width = 8,
    height = 6
)
```